---
title: "Toolbox"
output: pdf_document
toc: yes
---

# Introduction

This document presents the examples from Chapter 5 of Hadley Wickham's book **ggplot2**, "Toolbox".

This chapter presents a large number of examples for the display of a range of plot types, spanning many common circumstances of data presentation.

## Layering strategies

Wickham describes three purposes for a layer:

- **representing raw data**: we use pattern detection skills to identify gross and local structures, and outliers.
- **statistical summary of data**: it is useful to display models of data, and display predictions in the context of the data; this helps evaulate the model against the data.
- **addition of metadata**: the display of background context and annotations can help give meaning to raw data.

# Basic plot types

A handful of `geom`s constitute the basic building blocks of `ggplot2`. Most are associated with a named plot (i.e. when used on their own, the plot has a special name). Each of these `geom`s is two-dimensional, requiring `x` and `y` aesthetics, understanding `size` and `colour` aesthetics. Filled `geom`' understand `fill`, while the point `geom` uses shape, and the line and path `geom`s understand `linetype`.

- `geom_area()`: draws an **area plot**, a line plot filled to the *y*-axis
- `geom_bar(stat="identity")`: makes a **barchart**; `stat="identity"` is required as the default `stat` counts values.
- `geom_line()`: makes a **line plot**, where points are connected from left to right
- `geom_path()`: makes a **line plot**, where points are connected in the order they occur in the data
- `geom_point()`: renders a  **scatterplot**
- `geom_polygon()`: draws polygons, which are filled paths; each polygon vertex requires a separate row in the data
- `geom_text()`: adds labels at specified points; it requires an additional aesthetic, `label`
- `geom_tile()`: makes an **image plot**, where tiles are a regular tessellation of the plane

```{r ex5_1, fig.width=7, fig.height=6, message=F, warning=F}
library(ggplot2)
library(gridExtra)
df = data.frame(x=c(3, 1, 5), y=c(2, 4, 6),
                label=c("a", "b", "c"))
p = ggplot(df, aes(x, y, label=label)) + xlab(NULL) + ylab(NULL)
grid.arrange(p + geom_point() + labs(title="geom_point"),
             p + geom_bar(stat="identity") + labs(title="geom_bar"),
             p + geom_line() + labs(title="geom_line"),
             p + geom_area() + labs(title="geom_area"),
             p + geom_path() + labs(title="geom_path"),
             p + geom_text() + labs(title="geom_text"),
             p + geom_tile() + labs(title="geom_tile"),
             p + geom_polygon() + labs(title="geom_polygon"),
             ncol=4)
```

# Displaying distributions

Several `geom`s are available to display distributions, depending on how many dimensions the distribution has, whether it's continuous or discrete, or whether you want conditional or joint distributions.

For 1D continuous distributions, the **histogram** is the most important `geom`. It is always important to experiment with the `binwidth` and/or `breaks` options to obtain a suitable view onto the data that reveals the relevant characteristics. Never rely on the default values to give you insight into your data. 

```{r ex5_2, fig.width=7, fig.height=3, message=F, warning=F}
p = ggplot(diamonds, aes(depth)) + xlim(55, 70)
grid.arrange(p + geom_histogram(aes(y=..density..)),
             p + geom_histogram(aes(y=..density..), binwidth=0.1),
             ncol=2)
```

To compare between groups: small multiples plots can be used (`facets = . ~ var`); a frequency polygon (`geom="freqpoly"`) can be used; or a conditional density plot may be suitable (`position="fill"`), as illustrated below.

```{r ex5_3, fig.width=5, fig.height=4, message=F, warning=F}
p = ggplot(diamonds, aes(depth)) + xlim(58, 68)
p + geom_histogram(aes(y=..density..), binwidth=0.1) + facet_grid(cut ~ .)
p + geom_histogram(aes(fill=cut), binwidth=0.1, position="fill")
p + geom_freqpoly(aes(y=..density.., colour=cut), binwidth=0.1)
```

Both histogram and frequency polygon plots use `stat_bin`, which gives `..count..` (default) and `..density..` output variables. The `..density..` variable, being normalised to the total count, is useful for comparison of distribution shapes.

**Box-and-whisker** plots are used for continuous variables conditionsed by a categorical variable. The example below show boxplots conditioned on categorical and continuous variables.

```{r ex5_4, fig.width=5, fig.height=3, message=F, warning=F}
# We need plyr for round_any() below. round_any() rounds the first argument 
# number to a multiple of the second argument; floor means that the first
# argument rounds down, rather than up (the default).
library(plyr)
qplot(cut, depth, data=diamonds, geom="boxplot")
qplot(carat, depth, data=diamonds, geom="boxplot", 
      group=round_any(carat, 0.1, floor), xlim=c(0, 3))
```

**Jitter** plots are a crude way of assessing discrete distributions by adding random noise to reduce overplotting.

```{r ex5_5, fig.width=5, fig.height=3, message=F, warning=F}
qplot(class, cty, data=mpg, geom="jitter")
qplot(class, drv, data=mpg, geom="jitter")
```

**Density** plots are a smoothed variant of frequency polygons, where the curve is smoothed by a kernel. These are appropriate when the underlying density is smooth, continuous, and unbounded. The `adjust` parameter varies the smoothness of the density.

```{r ex5_6, fig.width=5, fig.height=3, message=F, warning=F}
qplot(depth, data=diamonds, geom="density", xlim=c(54, 70))
qplot(depth, data=diamonds, geom="density", xlim=c(54, 70), fill=cut, alpha=I(0.2))
```

# Dealing with overplotting

When large datasets are used, scatterplots can be difficult to interpret due to overplotting. If there are too many points, relationships can be obscured as it is difficult to see the density of data in any particular area and, in the worst case, all that can be seen is the extent of the data.

To reduce the effect of small amounts of overplotting, smaller and hollow glyphs may be used:

```{r ex5_7, fig.width=7, fig.height=3, message=F, warning=F}
# Two 2000 point samples
df = data.frame(x=rnorm(2000), y=rnorm(2000))
norm = ggplot(df, aes(x, y))
grid.arrange(norm + geom_point() + labs(title="overplot"),
             norm + geom_point(shape=1) + labs(title="hollow"),
             norm + geom_point(shape=".") + labs(title="small"),
             ncol = 3)
```

If the dataset is larger, it may be worth reducing glyph transparency with `alpha`. Specifying `alpha` as a ratio, e.g. as `1/n`, means that `n` points need to be overplotted to give a solid colour; the smallest allowable value is 1/256. 

```{r ex5_8, fig.width=7, fig.height=3, message=F, warning=F}
# Same examples as before, but we need the scales library to use alpha()
library(scales)
grid.arrange(norm + geom_point(colour=alpha("black", 1/3)) + labs(title="1/3"),
             norm + geom_point(colour=alpha("black", 1/5)) + labs(title="1/5"),
             norm + geom_point(colour=alpha("black", 1/10)) + labs(title="1/10"),
             ncol = 3)
```

For discrete data, jittering may be useful, especially in conjunction with transparency. The default jitter is 40% of the data resolution, to leave a small gap between adjacent regions. 

```{r ex5_9, fig.width=7, fig.height=6, message=F, warning=F}
td = ggplot(diamonds, aes(table, depth)) + xlim(50, 70) + ylim(50, 70)
jit = position_jitter(width=0.5)   # Set jitter to 0.5 (absolute value)
grid.arrange(td + geom_point() + labs(title="geom_point"),
             td + geom_jitter() + labs(title="geom_jitter"),
             td + geom_jitter(position=jit) + labs(title="geom_jitter"),
             td + geom_jitter(position=jit, colour=alpha("black", 1/10)) + labs(title="geom_jitter, 1/10"),
             td + geom_jitter(position=jit, colour=alpha("black", 1/50)) + labs(title="geom_jitter, 1/50"),
             td + geom_jitter(position=jit, colour=alpha("black", 1/200)) + labs(title="geom_jitter, 1/200"),
             ncol = 3)
```

The main problem of overplotting is that it is difficult to see the density of data in any particular area, so we might treat the issue as a 2D density estimation, instead. The brute force approach is to bin the data in 2D, and visualise the count as the 2D generalisation of a histogram. The usual way to do this is to use squares but this can lead to visual artefacts, and in these instances a hexagonal binning strategy with `geom_hexagon` or `stat_binhex` can be useful, for which the `hexbin` library is required. As ever with binned counts, it is wise never to accept the defaults without experimentation to determine which views give the most informative output.

```{r ex5_10, fig.width=7, fig.height=6, message=F, warning=F}
# To use stat_binhex() we need the hexbin library
library(hexbin)
d = ggplot(diamonds, aes(carat, price)) + xlim(1, 3) + theme(legend.position="none")
grid.arrange(d + stat_bin2d() + labs(title="stat_bin2d"),
             d + stat_bin2d(bins=10) + labs(title="bins=10"),
             d + stat_bin2d(binwidth=c(0.02, 200)) + labs(title="c(0.02, 200)"),
             d + stat_binhex() + labs(title="stat_binhex"),
             d + stat_binhex(bins=10) + labs(title="bins=10"),
             d + stat_binhex(binwidth=c(0.02, 200)) + labs(title="binwidth=c(0.02, 200)"),
             ncol = 3)
```

Another approach here is to estimate the underlying 2D density using `stat_density2d`. Once this is done, then the density contours can be overlaid, the density shown as coloured tiles, or points with size proportinal to density, and so on.

```{r ex5_11, fig.width=7, fig.height=6, message=F, warning=F}
d = ggplot(diamonds, aes(carat, price)) + xlim(1, 3) + theme(legend.position="none")
grid.arrange(d + geom_point() + geom_density2d() + labs(title="contours"),
             d + stat_density2d(geom="point", aes(size=..density..), contour=F) + labs(title="size proportional"),
             d + stat_density2d(geom="tile", aes(fill=..density..), contour=F) + labs(title="image"),
             last_plot() + scale_fill_gradient(limits=c(1e-5, 8e-4)) + labs(title="contour restriction"),
             ncol = 2)
```

# Drawing maps

The `maps` package provides some maps, and tools for map display. These can be incorporated into `ggplot2` output so you can, for example, add reference outlines to plots of spatial data, or to construct a chloropleth map. Map borders are added with the `borders()` function; the first two arguments indicate the `map` and the `region` within the map to be displayed. The rest of the arguments control the appearance of borders with `colour`, `size` and `fill`.

```{r ex5_12, fig.width=7, fig.height=3, message=F, warning=F}
library(maps)
data(us.cities)
big_cities = subset(us.cities, pop>5e5)
tx_cities = subset(us.cities, country.etc=="TX")
grid.arrange(qplot(long, lat, data=big_cities) + borders("state", size=0.5) + labs(title="large US cities"),
             ggplot(tx_cities, aes(long, lat)) + borders("county", "texas", colour="grey70") + geom_point(colour=alpha("black", 0.5)) + labs(title="Texas cities"),
             ncol=2)
```

It is more difficult to render chloropleth maps, in part because data identifiers need to match map data identifiers. The example below uses `map_data()` to convert a map into a data frame, which is then `merge()`d with data to produce a chloropleth map.

```{r ex5_13, fig.width=7, fig.height=6, message=F, warning=F}
states = map_data("state")
arrests = USArrests
names(arrests) = tolower(names(arrests))
arrests$region = tolower(rownames(USArrests))
choro = merge(states, arrests, by="region")
# Reorder rows so that the original polygon ordering is 
# retained.
choro = choro[order(choro$order),]
grid.arrange(qplot(long, lat, data=choro, group=group, fill=assault, geom="polygon") + labs(title="assault"),
             qplot(long, lat, data=choro, group=group, fill=assault/murder, geom="polygon") + labs(title="assault/murder"),
             ncol=1)
```

The `map_data()` function is useful for processing map data.

```{r ex5_14, fig.width=7, fig.height=4, message=F, warning=F}
ia = map_data("county", "iowa")
mid_range = function(x) mean(range(x, na.rm=TRUE))  # function to find the midpoint of a range of values
# ddply() breaks a dataframe up into subsets on the basis of some value,
# applys a function to each subset, and joins the results back into a dataframe
centres = ddply(ia, .(subregion), colwise(mid_range, .(lat, long)))  # Centre point of each county
ggplot(ia, aes(long, lat)) + geom_polygon(aes(group=group), fill=NA, colour="grey60") + geom_text(aes(label=subregion), data=centres, size=2, angle=45)
```