---
title: '`ggplot2` Grammar'
output: pdf_document
toc: yes
---

# Introduction

This document presents the examples from Chapter 3 of Hadley Wickham's book **ggplot2**, "Mastering the grammar".

The layered grammar of `ggplot2` is based on Wilkinson's "Grammar of Graphics." The grammar is useful because it allows the user to iteratively update plots one feature at a time. The modular design also enables developers to add a single component (such as a statistical transformation) at a time, and use the existing grammar components i conjunction with it.

# Dataset

We will be using the `mpg` dataset provided by `ggplot2`

```{r imports}
library(ggplot2)
head(mpg)
```

# Using `qplot()`

We can use `qplot()` to simply render a scatterplot of two continuous variables (engine displacement and mpg), colouring points by number of cylinders:

```{r qplot, fig.width=5, fig.height=3}
qplot(displ, hwy, data=mpg, colour=factor(cyl))
```

# A more complex plot

This plot adds facets, multiple layers and statistics to the basic concepts of the grammar:

```{r complex, fig.width=5, fig.height=3, message=F, warning=F}
qplot(displ, hwy, data=mpg, facets=. ~ year) + geom_smooth()
```

# Data structures

The following code illustrates the `ggplot2` data structure

```{r datastructure}
p = qplot(displ, hwy, data=mpg, colour=factor(cyl))
summary(p)
```