---
title: "Toolbox"
output: pdf_document
toc: yes
---

# Introduction

This document presents the examples from Chapter 5 of Hadley Wickham's book **ggplot2**, "Toolbox".

This chapter presents a large number of examples for the display of a range of plot types, spanning many common circumstances of data presentation.

## Layering strategies

Wickham describes three purposes for a layer:

- **representing raw data**: we use pattern detection skills to identify gross and local structures, and outliers.
- **statistical summary of data**: it is useful to display models of data, and display predictions in the context of the data; this helps evaulate the model against the data.
- **addition of metadata**: the display of background context and annotations can help give meaning to raw data.

# Basic plot types

A handful of `geom`s constitute the basic building blocks of `ggplot2`. Most are associated with a named plot (i.e. when used on their own, the plot has a special name). Each of these `geom`s is two-dimensional, requiring `x` and `y` aesthetics, understanding `size` and `colour` aesthetics. Filled `geom`' understand `fill`, while the point `geom` uses shape, and the line and path `geom`s understand `linetype`.

- `geom_area()`: draws an **area plot**, a line plot filled to the *y*-axis
- `geom_bar(stat="identity")`: makes a **barchart**; `stat="identity"` is required as the default `stat` counts values.
- `geom_line()`: makes a **line plot**, where points are connected from left to right
- `geom_path()`: makes a **line plot**, where points are connected in the order they occur in the data
- `geom_point()`: renders a  **scatterplot**
- `geom_polygon()`: draws polygons, which are filled paths; each polygon vertex requires a separate row in the data
- `geom_text()`: adds labels at specified points; it requires an additional aesthetic, `label`
- `geom_tile()`: makes an **image plot**, where tiles are a regular tessellation of the plane

```{r ex5_1, fig.width=7, fig.height=6, message=F, warning=F}
library(ggplot2)
library(gridExtra)
df = data.frame(x=c(3, 1, 5), y=c(2, 4, 6),
                label=c("a", "b", "c"))
p = ggplot(df, aes(x, y, label=label)) + xlab(NULL) + ylab(NULL)
grid.arrange(p + geom_point() + labs(title="geom_point"),
             p + geom_bar(stat="identity") + labs(title="geom_bar"),
             p + geom_line() + labs(title="geom_line"),
             p + geom_area() + labs(title="geom_area"),
             p + geom_path() + labs(title="geom_path"),
             p + geom_text() + labs(title="geom_text"),
             p + geom_tile() + labs(title="geom_tile"),
             p + geom_polygon() + labs(title="geom_polygon"),
             ncol=4)
```

# Displaying distributions

Several `geom`s are available to display distributions, depending on how many dimensions the distribution has, whether it's continuous or discrete, or whether you want conditional or joint distributions.

For 1D continuous distributions, the **histogram** is the most important `geom`. It is always important to experiment with the `binwidth` and/or `breaks` options to obtain a suitable view onto the data that reveals the relevant characteristics. Never rely on the default values to give you insight into your data. 

```{r ex5_2, fig.width=7, fig.height=3, message=F, warning=F}
p = ggplot(diamonds, aes(depth)) + xlim(55, 70)
grid.arrange(p + geom_histogram(aes(y=..density..)),
             p + geom_histogram(aes(y=..density..), binwidth=0.1),
             ncol=2)
```

To compare between groups: small multiples plots can be used (`facets = . ~ var`); a frequency polygon (`geom="freqpoly"`) can be used; or a conditional density plot may be suitable (`position="fill"`), as illustrated below.

```{r ex5_3, fig.width=5, fig.height=4, message=F, warning=F}
p = ggplot(diamonds, aes(depth)) + xlim(58, 68)
p + geom_histogram(aes(y=..density..), binwidth=0.1) + facet_grid(cut ~ .)
p + geom_histogram(aes(fill=cut), binwidth=0.1, position="fill")
p + geom_freqpoly(aes(y=..density.., colour=cut), binwidth=0.1)
```

Both histogram and frequency polygon plots use `stat_bin`, which gives `..count..` (default) and `..density..` output variables. The `..density..` variable, being normalised to the total count, is useful for comparison of distribution shapes.

**Box-and-whisker** plots are used for continuous variables conditionsed by a categorical variable. The example below show boxplots conditioned on categorical and continuous variables.

```{r ex5_4, fig.width=7, fig.height=3, message=F, warning=F}
# We need plyr for round_any() below. round_any() rounds the first argument 
# number to a multiple of the second argument; floor means that the first
# argument rounds down, rather than up (the default).
library(plyr)
qplot(cut, depth, data=diamonds, geom="boxplot")
qplot(carat, depth, data=diamonds, geom="boxplot", 
      group=round_any(carat, 0.1, floor), xlim=c(0, 3))
```